A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-12 example.tex
More step by step example

\begin{figure}[h]
    \centering 
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a_annotated}
    \caption{ $A'$ with its associated transition variables in symbolic form.
    Note: unused ($0$) self-loops at $I_{A'}$ and $F_{A'}$ have been pruned for
    readability.}\label{fig:aut_a_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b_annotated}
    \caption{$B$ with its associated transition variables in symbolic forms.
    Note that state $B$ is now completely dead.}\label{fig:aut_b_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The starting automata, without their counter increments and with
  their transition variables in symbolic form.}\label{fig:propagated}
\end{figure}

@@

Performing symbolic reasoning on these equations and plugging the corresponding
symbolic representation of the constraints on each corresponding integer
variable back as an annotation on its transition, we obtain the two automata in
\cref{fig:propagated}.

Several transitions are already marked as unuseable (have a transition variable
value of $0$) or always present (have a transition variable with a constant
value greater than zero) by just applying standard linear equation solving to
the flow equations described above. Note that this includes
\cref{const:more-inside-than-before,const:something-before-and-after}, which are
related to the transition variables through the sums described in
\cref{eq:counter-sums}. This means that we effectively can prune $A'$ of its
upper path when continuing the proof, as it is incongruent with the flow
equations and \cref{ex:string-constraints}. Note that the per-automata length
counting registers have been assigned the same solver variable $l_A$. Since they
have to be equal, either one can be used in both automata.

Notably, one such conclusion is that $i > 0$ from
\cref{const:something-before-and-after}. From flow preservation we have that the
$F_{A'}$ to $F_{A'}$ transition is represented by $l_A - n - i$.
\footnote{Intuitively, the total length is made up of what is before, inside and
after the substring.} This in turn through \cref{const:more-inside-than-before}
forces $n \geq 2$, which in turn (through flow-preservation) forces a path from
$I_{A'}$ to $F_{A'}$ through $D$, rather than the single-transition path from
$I_{A'}$ to $F_{A'}$. This immediately implies that the problem is
unsatisfiable, since we know this transition has no correspondance in automaton
$B$.

However, \Calculus{} does not at this point, and we press on with concluding
that none of the loops of $A'$ at this stage can become disconnected since
they are at the initial and final states, both of which are trivially reachable
for any assignments of the non-constant transition variables. This means that
all valid counter assignments of $A'$ are described by \cref{eq:counter-sums}
and the flow equations alone without the need for expensive connectivity
constraints.

>>>>

@@

\begin{figure}[h]
  \centering 
  \includegraphics[scale=\autscale]{a_1}
  \caption{ $A'$ with its associated transition variables in symbolic form after
  minimal elminiation of variables using linear algebra. Note the incoming flow
  of $1$ in the starting state and its balancing of the outgoing transitions!
  }\label{fig:a_1}
\end{figure}

As an example, we start with reasoning about $A'$. Performing symbolic reasoning
on these equations and plugging the corresponding symbolic representation of the
constraints on each corresponding integer variable back as an annotation on its
transition, we obtain the automaton in \cref{fig:a_1}. After slightly more
linear algebra to remove variables through equality elimination, we have
\cref{fig:a_2}, whose transitions are only described in terms of the free
variables we care about representing the length of the automaton ($l_A$), the
start of the substring ($i$), and the length of the substring ($n$).

\begin{figure}[h]
  \centering 
  \includegraphics[scale=\autscale]{a_2}
  \caption{ $A'$ after even more linear algebra, with most transitions expressed
  in terms of $n$. Intuitively, this version captures the fact that $n$ is given
  by distributing the incoming flow of $1$ across the two outgoing transitions
  from $I_{A'}$, the initial state.}
  \label{fig:a_2}
\end{figure}

Having obtained this representation, we conclude that $1 < n \leq 2$ from
\cref{const:something-before-and-after,const:something-before-and-after}. The
upper bound is obtained by the following reasoning. Since all transition
variables are nonnegative integers (a transition cannot be used a negative
amount of times), $n \leq 2$ from the $I_A$ to $F_A$ transition. Therefore, it
follows that $n=2$, which implies that the $I_A$ to $F_A$ transition can never
be used under these constraints and that we must use the two $\Char{d}$-labelled
transitions, both now taken $n-1 =1$ times. This immediately implies that the problem is
unsatisfiable, since we know this transition has no correspondance in automaton
$B$.

We repeat similar methods to arrive at the automata in \cref{fig:propagated}.

\begin{figure}[h]
    \centering 
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a_annotated}
    \caption{ $A'$ with its associated transition variables in symbolic
    form.}\label{fig:aut_a_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b_annotated}
    \caption{$B$ with its associated transition variables in symbolic forms.
    Note the large number of implicitly existence-quantified bound variables on
    transitions, suggesting that this automaton has a more complex structure
    with relation to its (free) target variable representing the string length
    than $A'$. The variables used here are fresh, and do not correspond to the
    ones previously used.}\label{fig:aut_b_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The automata after applying linear algebra and equality elimination
  to prune them.}\label{fig:propagated}
\end{figure}

Note that the per-automata length counting registers have been assigned the same
solver variable $l_A$. Since they have to be equal, either one can be used in
both automata through similar applications of equality elimination rules.

At this point we know we have an unsatisfiable solution, \Calculus{} does not.
We press on with concluding that none of the loops of $A'$ at this stage can
become disconnected since they are at the initial and final states, both of
which are trivially reachable for any assignments of the non-constant transition
variables. This means that all valid counter assignments of $A'$ are described
by \cref{eq:counter-sums} and the flow equations alone without the need for
expensive connectivity constraints.
